# Analysis of longitudinal data

## RATS data
Read in the data and convert the `ID`and `Group` columns to factors.
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
RATSL <- read.table("data/RATSL.txt", header = TRUE)
RATSL$ID <- as.factor(RATSL$ID)
RATSL$Group <- as.factor(RATSL$Group)
glimpse(RATSL)
```
The data seems to be in order.  
Next to the analyses. 

### Non-standardized data
```{r, fig.width=10}
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

### Standardization
As higher weight in the beginning seems to mean also higher weight in the end, standardization is required.  
$$\ standardized(x) = \frac{x-mean(x)}{sd(x)}$$

```{r, fig.width=10}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight-mean(Weight))/sd(Weight)) %>%
  ungroup()

ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$stdweight), max(RATSL$stdweight)), name = "standardized bprs")
```

### Mean response for treatments

The starndard error of the mean can be calculated with the following equation:
$$\ se = \frac{sd(x)}{\sqrt{n}} $$

```{r}
n <- RATSL$Time %>% unique() %>% length()
RATS_trmt <- RATSL %>% group_by(Group, Time) %>% summarize(mean = mean(Weight), se = sd(Weight)/sqrt(n))
glimpse(RATS_trmt)

ggplot(RATS_trmt, aes(x=Time, y=mean, linetype = Group, shape = Group)) + 
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=2) +
  scale_shape_manual(values = c(1,2,5)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.5) +
  theme(legend.position = c(0.8,0.5), panel.grid = element_blank(), panel.background = element_blank(), panel.border = element_rect(fill = FALSE), legend.key = element_rect(fill=FALSE)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)") 

ggplot(RATSL, aes(x= as.factor(Time), y=Weight, fill=Group)) +
  geom_boxplot() +
  theme(legend.position = c(0.8,0.4), panel.grid = element_blank(), panel.background = element_blank(), panel.border = element_rect(fill = FALSE), legend.key = element_rect(fill=FALSE)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)") + scale_x_discrete(name = "Time") + 
  scale_fill_manual(values=c("black", "grey50", "white"))
```

### Outlier in the data

```{r}
RATSL8 <- RATSL %>% 
  group_by(Group, ID) %>%
  summarize(mean = mean(Weight)) %>%
  ungroup()

ggplot(RATSL8, aes(x=Group, y=mean)) + 
  geom_boxplot() + 
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") + 
  theme(legend.position = c(0.8,0.8), panel.grid = element_blank(), panel.background = element_blank(), panel.border = element_rect(fill = FALSE)) +
  scale_y_continuous(name = "mean(Weight) per group")
```

There's one outlier in the group 2. It will be removed.
One outlier data point can be removed. And the data plotted again.

### Without the outlier
```{r}
RATSL8S1 <- filter(RATSL8, (Group=="2" & mean<500) | Group!="2")
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  theme(legend.position = c(0.8,0.8), panel.grid = element_blank(), panel.background = element_blank(), panel.border = element_rect(fill = FALSE)) +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight) per group")
```

### Test group differences
Seems obvious that there's differences between groups. But it mainly comes from the starting weight. 
We can still test the difference between groups. 

```{r}
RATS_lm <- aov(mean ~ Group, data = RATSL8S1)
summary(RATS_lm)
```

As expected there was a difference between groups. We can do a *post hoc* test to see pairwise differences. 

```{r}
TukeyHSD(RATS_lm)
```

## BPRS data
Read in the data and convert the `treatment`and `subject` columns to factors.
```{r}
BPRSL <- read.table("data/BPRSL.txt", header = TRUE)
BPRSL$treatment <- as.factor(BPRSL$treatment)
BPRSL$subject <- as.factor(BPRSL$subject)
glimpse(BPRSL)
```
The data seems to be in order.  
Next to the analyses. 

### Data overlook

```{r}
ggplot(BPRSL, aes(x=week, y=bprs, group=subject, shape=treatment)) + 
  geom_point() +
  theme(legend.position = "top", panel.grid = element_blank(), panel.background = element_blank(),
        panel.border = element_rect(fill = FALSE), legend.key = element_rect(fill=FALSE))

ggplot(BPRSL, aes(x = week, y = bprs, fill=subject)) +
  geom_line(aes(linetype = treatment)) +
  theme(legend.position = c(0.8,0.8), panel.grid = element_blank(), panel.background = element_blank(),
        panel.border = element_rect(fill = FALSE), legend.key = element_rect(fill=FALSE)) + 
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 2), expand=c(0,0)) + scale_y_continuous(name = "bprss") + theme(legend.position = "top")
```

### Linear mixed effects models for repeated measurements

First we fit a random intercept model
```{r, message=FALSE}
library(lme4)
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
```
Then we fit a random intercept and random slope model and test wheter it's better than the random intercept only model.
```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)
anova(BPRS_ref, BPRS_ref1)
```

It was better. So last we can try an interaction between week and treatment.
```{r}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)
anova(BPRS_ref2, BPRS_ref1)
```
That was not better, so we'll stick with the model witout the interaction and plot that.

```{r}
library(gridExtra)
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, fill = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 2)) +
  scale_y_continuous(name = "Observed bprs") +
  theme(panel.grid = element_blank(), panel.background = element_blank(),
        panel.border = element_rect(fill = FALSE), legend.key = element_rect(fill=FALSE), 
        legend.position = "top")

Fitted <- fitted(BPRS_ref1)
BPRSL <- mutate(BPRSL, fitted=Fitted)

p2 <- ggplot(BPRSL, aes(x = week, y = fitted, fill = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 2)) +
  scale_y_continuous(name = "Fitted bprs") +
  theme(panel.grid = element_blank(), panel.background = element_blank(),
        panel.border = element_rect(fill = FALSE), legend.key = element_rect(fill=FALSE), 
        legend.position = "top")
grid.arrange(p1, p2, ncol=2)
```

**That's it.**